---
title: "Modelling Zoonoses in Rodent Communities"
author: "David Redding, Greg Milne, Ana Martinez-Checa, Harry Gordon"
date: "07/06/2024"
bibliography: 01_references.bib
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is the working manuscript for a study on disease modelling in rodent communities by the Infectious Disease Systems Ecology group at the British Natural History Museum.

David Simons' paper: @simons2023.

Paper on causal inference: @dee2023.

## Methods

These are the methods.

Here is a world map displaying locations of our sampled rodents:

```{r plot_world_map, echo=FALSE}
# Load necessary libraries
library(ggplot2)
library(maps)
library(mapdata)

# Load data frame
combined_data <- readRDS("../data/combined_data.rds")
attach(combined_data)
# Get the world map data
world <- map_data("world")

# Plot the points onto the world map 
ggplot() +
  geom_polygon(data = world, aes(x = long, y = lat, group = group), fill = "lightgray", color = "darkgray") +
  geom_point(data = host, aes(x = decimalLongitude, y = decimalLatitude), color = "red", size = 2) +
  theme_minimal() +
  coord_fixed(1.3) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank())

```

## Results

These are the results.

## Discussion

This is the discussion.

## References