---
title: "Modelling Zoonoses in Rodent Communities"
author: "David Redding, Greg Milne, Ana Martinez-Checa, Harry Gordon"
date: "07/06/2024"
output: pdf_document
bibliography: 01_references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is the working manuscript for a study on disease modelling in rodent communities by the Infectious Disease Systems Ecology group at the British Natural History Museum.

David Simons' paper: @simons_rodent_2023.

GLC: @zhang_glc_fcs30d_2024

## Methods

These are the methods.

Here is a world map displaying locations of our sampled rodents:

```{r plot_world_map, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# Load data frame
host_path <- readRDS("../data/host_path_wide.rds")
attach(host_path)
# Get the world map data

world <- ne_countries(scale = "large", returnclass = "sf")
coords_sf <- st_as_sf(host_path, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326, agr = "constant")
host_path$year <- as.numeric(format(host_path$start_date, "%Y"))
# Plot the points onto the world map 

ggplot(data = world) +
  geom_sf(colour="darkgray") +
  geom_sf(data = coords_sf, aes(geometry = geometry, colour=host_path$year), size = 2, shape=18) +
  coord_sf(expand = FALSE) +                    # Keep the aspect ratio
  labs(title = "Locations of Trapped Rodents") + 
  theme_void() +
  theme(panel.background = element_rect(), 
        axis.text = element_blank(),            # Remove axis text (tick labels)
        axis.ticks = element_blank()) +
  scale_color_gradient(low = "#FF9D00", high = "#FF0000", name = "Year")
  
```

```{r plot_timeline, include=FALSE, echo=FALSE, fig.height = 2, message=FALSE, error=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)

# plot the study dates on a timeline
ggplot(host_path, aes(x = start_date, y = 1)) +
  geom_segment(aes(x = start_date, xend = start_date, y = 0.9, yend = 1.1), color = "red") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_continuous(limits = c(0.5, 1.5)) +
  labs(title = "Timeline of Samples", x = "Date", y = "") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())

```

Here is the infectivity rates of our sampled rodents:

```{r plot_infection, echo=FALSE, message=FALSE, warning=FALSE, fig.width=40, fig.height=25}
pacman::p_load(here,
               stringr,
               dplyr,
               ape,    # for phylogenetic analyes
               ggtree, # for plotting trees
               ggplot2,
               tidyr)
host.pathogen.data <- readRDS(here("data", "host_path_wide.rds"))

# Correct species names
mutated.host.pathogen.data <- host.pathogen.data %>%
  mutate(associatedTaxa = case_when(
    associatedTaxa == "Akodon azare" ~ "Akodon azarae",
    associatedTaxa == "brush mice" ~ "Peromyscus boylii",
    associatedTaxa == "Clethrionomys glareolus" ~ "Myodes glareolus",
    associatedTaxa == "Clethryonomys glareolus" ~ "Myodes glareolus",
    associatedTaxa == "Oligoryzomys favescens" ~ "Oligoryzomys flavescens",
    associatedTaxa == "P. californicus" ~ "Peromyscus californicus",
    associatedTaxa == "dusky-footed" ~ "Neotoma fuscipes",
    TRUE ~ associatedTaxa  # Retain original name if no match
  ))

# Ensure 'positive' and 'negative' are numeric
plot.positives.negatives <- mutated.host.pathogen.data %>%
  mutate(positive = as.numeric(positive),
         negative = as.numeric(negative),
         individualCount = as.numeric(individualCount))

# Aggregate the data by associatedTaxa
plot_aggregated <- plot.positives.negatives %>%
  group_by(associatedTaxa) %>%
  summarize(
    positives = sum(positive, na.rm = TRUE),
    negatives = sum(negative, na.rm = TRUE),
    abundance = sum(individualCount, na.rm = TRUE),
  ) %>%
  ungroup()

# Calculate the % of positives
aggregated_data1 <- plot_aggregated %>%
  mutate(positives = ifelse((positives + negatives) == 0, NA, 
                                   (positives / (positives + negatives)) * 100))

# Remove rows with NA values for % positive
aggregated_data2 <- aggregated_data1 %>%
  filter(!is.na(positives))

# Calculate the % of negatives
aggregated_data3 <- aggregated_data2 %>%
  mutate(negatives = ifelse((positives + negatives) == 0, NA, 
                                   (negatives / (positives + negatives)) * 100))

# Remove rows with NA values for % positive
aggregated_data4 <- aggregated_data3 %>%
  filter(!is.na(negatives))

# Sort data by abundance from less to more
aggregated_data5 <- aggregated_data4 %>%
  arrange(abundance)

# Aggregate in % of infectivity
plot.infectivity <- aggregated_data5 %>%
  pivot_longer(cols = c("positives", "negatives"), 
               names_to = "Result", 
               values_to = "Results")
# Plot
ggplot(plot.infectivity, aes(x = associatedTaxa, y = abundance, fill = Result)) +
  geom_col(position = "stack", width = 0.7) +
  scale_fill_manual(values = c("positives" = "orange", "negatives" = "olivedrab3")) +
  labs(x = "Associated Taxa", y = "abundance", fill = "Test Result in %") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r plot_infection2, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10}
pacman::p_load(here,
               stringr,
               dplyr,
               ggplot2,
               tidyr)
host.pathogen.data <- readRDS(here("data", "host_path_wide.rds"))


# Correct species names
mutated.host.pathogen.data <- host.pathogen.data %>%
  mutate(associatedTaxa = case_when(
    associatedTaxa == "Akodon azare" ~ "Akodon azarae",
    associatedTaxa == "brush mice" ~ "Peromyscus boylii",
    associatedTaxa == "Clethrionomys glareolus" ~ "Myodes glareolus",
    associatedTaxa == "Clethryonomys glareolus" ~ "Myodes glareolus",
    associatedTaxa == "Oligoryzomys favescens" ~ "Oligoryzomys flavescens",
    associatedTaxa == "P. californicus" ~ "Peromyscus californicus",
    associatedTaxa == "dusky-footed" ~ "Neotoma fuscipes",
    TRUE ~ associatedTaxa  # Retain original name if no match
  ))

# Ensure 'positive' and 'negative' are numeric
plot.positives.negatives <- mutated.host.pathogen.data %>%
  mutate(positive = as.numeric(positive),
         negative = as.numeric(negative),
         individualCount = as.numeric(individualCount))

# Aggregate the data by associatedTaxa
plot_aggregated <- plot.positives.negatives %>%
  group_by(associatedTaxa) %>%
  summarize(
    positives = as.numeric(sum(positive, na.rm = TRUE)),
    negatives = as.numeric(sum(negative, na.rm = TRUE)),
    abundance = as.numeric(sum(individualCount, na.rm = TRUE)),
  ) %>%
  ungroup()

# Calculate the % of positives
aggregated_data1 <- plot_aggregated %>%
  mutate(percent_positive = ifelse((positives + negatives) == 0, NA, 
                                   (positives / (positives + negatives)) * 100))

# Remove rows with NA values for % positive
aggregated_data2 <- aggregated_data1 %>%
  filter(!is.na(percent_positive))

# Sort data by abundance from less to more
aggregated_data3 <- aggregated_data2 %>%
  arrange(abundance)

# Plot scatter graph
ggplot(aggregated_data3, aes(x = abundance, y = percent_positive, label = associatedTaxa, color = associatedTaxa)) +
  geom_point() +
  geom_text(size = 2, vjust = -1, hjust = 0.5) +  # Adjust text size
  labs(
    title = "Abundance vs % of Positives",
    x = "Abundance",
    y = "% of Positives"
  ) +
  theme_minimal()+
  theme(legend.position = "none")

```

```{r plot_pathogen, echo=FALSE, message=FALSE, warning=FALSE, fig.width=40, fig.height=25}
pacman::p_load(here,
               stringr,
               dplyr,
               ggplot2,
               tidyr)
host.pathogen.data <- readRDS(here("data", "host_path_wide.rds"))

# Correct pathogen names
mutated.pathogen.data <- host.pathogen.data %>%
  mutate(scientificName.y = case_when(
    scientificName.y == "SNV" ~ "Sin Nombre Virus",
    scientificName.y == "SEOV" ~ "Sin Nombre Orthohantavirus",
    scientificName.y == "" ~ "NA",
    TRUE ~  scientificName.y  # Retain original name if no match
  ))

# Ensure 'positive' and 'negative' are numeric
plot.positives.negatives <- mutated.pathogen.data %>%
  mutate(positive = as.numeric(positive),
         negative = as.numeric(negative),
         individualCount = as.numeric(individualCount))

# Aggregate the data by pathogen and tests
plot_aggregated <- plot.positives.negatives %>%
  group_by(scientificName.y, assay) %>%
  summarize(
    positives = sum(positive, na.rm = TRUE),
    negatives = sum(negative, na.rm = TRUE),
    tests = sum(individualCount, na.rm = TRUE),
  ) %>%
  ungroup()

# Calculate the % of positives
aggregated_data1 <- plot_aggregated %>%
  mutate(percent_positive = ifelse((positives + negatives) == 0, NA, 
                                   (positives / (positives + negatives)) * 100))

# Remove rows with NA values for % positive
aggregated_data2 <- aggregated_data1 %>%
  filter(!is.na(percent_positive))

# Arrange by type of test
aggregated_data3 <- aggregated_data2 %>%
  arrange(assay)

# Plot 
ggplot(aggregated_data3, aes(x = percent_positive, y = scientificName.y, color = assay)) +
  geom_jitter(size = 4, width = 0.2, height = 0) +  # Adds points with some jitter to avoid overplotting
  scale_color_brewer(palette = "Set2") +  # Use a color palette from RColorBrewer for better aesthetics
  labs(title = "Percentage of Positive Tests by Pathogen and Test Type",
       x = "Percentage of Positive Tests",
       y = "Pathogen") +
  theme_minimal(base_size = 14) +  # Minimal theme for a clean look
  theme(
    axis.text.x = element_text(size = 12),  # Size x-axis labels
    axis.text.y = element_text(size = 12),  # Size y-axis labels
    axis.title.x = element_text(size = 14),  # Size x-axis title
    axis.title.y = element_text(size = 14),  # Size y-axis title
    legend.title = element_text(size = 14),  # Size legend title
    legend.text = element_text(size = 12)    # Size legend text
  )

```

### Rodent phylogeny

Here is the mammal tree with our host species higlighted

```{r plot_phylogeny, echo=FALSE, message=FALSE, warning=FALSE}
# load necessary libraries
library(ggtree)
library(ggplot2)
library(here)

# load data
host.clade <- read.tree(here("data", "./phylogenetics/host.mrca.tree"))
pruned.host.spp <- readRDS(here("data", "./phylogenetics/harmonised.host.sp.names.rds")) %>%
  as.data.frame()

# find our species of interest
host.clade$in.data <- ifelse(host.clade$tip.label %in% pruned.host.spp$tree.names, 
       "purple", NA)
host.clade$in.data <- as.factor(host.clade$in.data)

# Plot the phylogeny
ggtree(host.clade, size = 0.05) + 
  geom_tippoint(color = host.clade$in.data, size = 3) + 
  geom_treescale(x = 0, y = 2200, col = "midnightblue")

```

### GLC database

Here is an example of a raster from the GLC database with the site coordinate plotted and a buffer of radius 1 km.

```{r buffer_plot, width="0.3\\linewidth", include=TRUE, echo=FALSE, out.width = "400px", fig.align='center'}
knitr::include_graphics("../figures/bufferplot.pdf")
```

### Data papers

We collected data from the following papers for use in this analysis: @burns_novel_2018, @cabrera_deciphering_2023, @dietrich_small-scale_1997, @fernandes_co-circulation_2015, @fulhorst_bear_2002, @jameson_continued_2013, @kosasih_evidence_2011, @ledina_dinara-new_2002, @leduc_epidemiological_1986, @mauldin_assessment_2013, @milazzo_ecology_2013, @nemirov_isolation_1999, @zhang_glc_fcs30d_2024, @seijo_study_2003, @sheikh_ali_first_2014, @suarez_prevalence_2003, @williams_outbreak_1997, @williamson_continuing_2021.

## References
