---
title: "SpREAD: Spatio-temporal Rodent Community Ecology And Disease Database."
subtitle: Harry Gordon^1^ * , Ana Martinez-Checa^1^ * , Greg Milne^1^, David Redding^1^, David Simons^2^
author:
  - ^1^ Natural History Musuem, London
  - ^2^ Pennsylvania State University
  - (*) Contributed equally
output: pdf_document
geometry: margin=2cm
fontsize: 20pt
bibliography: 01_references.bib
abstract: |
  Rodents are reservoirs of zoonotic diseases, posing a significant threat to human health. Understanding rodent ecology and population dynamics is therefore important for mitigating human zoonotic hazards. Trapping studies have improved our understanding of zoonotic epidemiology, however, capturing known reservoirs to the exclusion of other rodent species limits inference on broader eco-epidemiological questions. Detailed information on rodent communities could be used to understand how global changes in habitat suitability and climate influence the dynamic ecosystems from which zoonotic diseases emerge and are maintained. Currently, there is no standard method for reporting rodent trapping data. This project aims to establish a high-resolution rodent community seroprevalence database by collating published data on rodent abundance and pathogen prevalence, and provide a standard for its reporting going forward. These data will facilitate robust inference on the eco-epidemiological predictors of rodent-borne disease, helping to shape our understanding of zoonotic disease risk in a changing world.
header-includes: 
  - \renewcommand{\and}{\\}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    fig.width   = 10,
    fig.height  = 5,
    fig.align   = 'center',
    echo        = FALSE,
    message     = FALSE,
    warning     = FALSE,
    error       = FALSE,
    strip.white = TRUE,
    warning     = FALSE,
    cache = TRUE,
    fig.pos="h")
```




# Introduction

# Methods

## Requirements for inclusion

The minimum requirements for data to be included in the SpREAD database are as follows:

1. **Date of trapping** (minimum resolution of a month)
1. **Duration of trapping** (or start and end date)
1. **Locality or coordinate location** of trapping site
1. **All captured rodents' abundances** are reported
1. **All individuals are tested** for at least one pathogen
1. **Type of test used and pathogen name** (at least family level e.g. hantavirus)

Desirable:

1. Indication of trapping method (e.g. grid area/number of traps/type of trap)
1. Indication of trapping effort (e.g. trap nights)


# Results

Through extraction of data from 27 published articles (so far), we have gathered data on 189 unique rodent communities across 5 continents (fig. 1). These are comprised of 10,955 rodents, of which 480 were positive for a pathogen of interest. The most common rodents found were of the *sigmodon* (cotton rats) and *peromyscus* (deer mice) genera (fig. 2). The most prevalent viruses were Sin Nombre Virus, Catarina Virus and Puumula Orthohantavirus, all belonging to the family *hantaviridae* (fig. 3).

# References

# Tables and Figures

```{r plot_world_map, echo=FALSE, fig.cap="The geographic locations of trapped rodents included in our study. The colour of points denotes the year sampled."}
# Load necessary libraries
library(ggplot2)
library(viridis)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# Load data frame
host_path <- readRDS("../data/host_path_wide.rds")
attach(host_path)

# Get the world map data
world <- ne_countries(scale = "large", returnclass = "sf")
coords_sf <- st_as_sf(host_path, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326, agr = "constant")
host_path$year <- as.numeric(format(host_path$start_date, "%Y"))

# Plot the points onto the world map 
ggplot(data = world) +
  geom_sf(colour="darkgrey") +
  geom_sf(data = coords_sf, aes(geometry = geometry, colour=host_path$year), size = 3, shape=19) +
  coord_sf(expand = FALSE) +                    # Keep the aspect ratio
  theme_void() +
  theme(panel.background = element_rect(), 
        axis.text = element_blank(),            # Remove axis text (tick labels)
        axis.ticks = element_blank(),
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  scale_colour_viridis(name="Year", direction=1, option="inferno", limits=c(1980, 2030))
  
```


```{r plot_n_hosts, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.cap="The number of hosts trapped for each genus", fig.align='center'}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(scico)

# group by host and sum count
host_plot <- host_path %>% group_by(genus) %>% summarise(n_hosts=sum(as.numeric(individualCount)))
host_plot <- host_plot[-1, ] %>% arrange(desc(n_hosts)) %>% slice(1:12)

# Plot barchart
ggplot(host_plot, aes(x = reorder(genus, -n_hosts), y = n_hosts, fill=genus)) +
  geom_bar(stat = "identity") + 
  annotate("point", x = 9, y = 1500, shape = 21, size = 50, fill = "chocolate4", color = "chocolate4") +
  annotate("text", x = 9, y = 1500, label = "10955 rodents\n total", size = 5) +
  labs(x = "Genus",
       y = "Number Trapped") +
  theme_minimal() +
  theme(plot.margin = margin(0, 0, 0, 0, "cm"),
        axis.text.x = element_text(angle = 45, hjust = 1, size=11),
        legend.title = element_blank()) +
  scale_fill_scico_d(palette = "tokyo")

  
```


```{r plot_n_infections, fig.pos="h", echo=FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.cap="The number of infections for each virus tested", fig.align='center'}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(scico)

host_path <- readRDS("../data/host_path_wide.rds")
attach(host_path)

# group by virus and sum infections
infection_plot <- host_path %>% group_by(pathogen_abbrev) %>% summarise(n_infections=sum(positive))
infection_plot <- infection_plot[-1, ] %>% arrange(desc(n_infections))

# Plot barchart
ggplot(infection_plot, aes(x = reorder(pathogen_abbrev, -n_infections), y = n_infections, fill=pathogen_abbrev)) +
  geom_bar(stat = "identity") + 
  annotate("point", x = 10.45, y = 70, shape = 21, size = 50, fill = "cadetblue", color = "cadetblue") +
  annotate("text", x = 10.45, y = 70, label = "480 positive\n rodents", size = 5) +
  labs(x = "Pathogen",
       y = "Number of Infections") +
  theme_minimal() +
  theme(plot.margin = margin(0, 0, 0, 0, "cm"),
        axis.text.x = element_text(size=11),
        legend.title = element_blank()) +
  scale_fill_scico_d(palette = "roma")

  
```


